---
title: "Module 1 — Session 2: From ASVs to biology"
subtitle: "27221 Microbiome Engineering — Metabarcoding of wastewater treatment plant communities"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Before you start

Last session you turned raw reads into an **ASV table**. Today you'll assign
those ASVs a taxonomy using the **MiDAS 4** reference database (the
purpose-built reference for wastewater treatment plant bacteria), attach your
sample metadata, and ask a real engineering question: *does the microbial
community track with how the plant is run?*

By the end of today you should have produced **one figure** and a short
written interpretation — that's your deliverable for this module.

---

## 0. Setup

```{r libraries}
library(dada2)
library(phyloseq)
library(tidyverse)
library(vegan)
```

```{r config}
# --- SESSION CONFIG: adjust if needed ---
output_dir      <- "~/module1_output"                 # same folder you saved to in Session 1
midas_ref_path  <- "/path/to/shared/wwtp_16S/MiDAS4.fasta"  # MiDAS 4 reference for assignTaxonomy()
```

```{r load-session1-output}
seqtab.nochim <- readRDS(file.path(output_dir, "seqtab_nochim.rds"))
metadata <- read_csv(file.path(output_dir, "metadata_used.csv"))
dim(seqtab.nochim)
```

---

## 1. Assign taxonomy against MiDAS 4

```{r assign-taxonomy}
taxa <- assignTaxonomy(seqtab.nochim, midas_ref_path, multithread = TRUE)
head(taxa)
```

**Question:** How far down the taxonomic ranks (domain → species) do most of
your ASVs get classified? MiDAS 4 was built specifically for WWTP bacteria —
if you were to instead use a generic universal reference database, would you
expect classification depth to improve, worsen, or stay the same, and why?

---

## 2. Build a phyloseq object

`phyloseq` is a standard R package for holding an ASV/OTU table, a taxonomy
table, and sample metadata together, and for computing common microbiome
statistics/plots on them.

```{r build-phyloseq}
metadata_df <- metadata %>% column_to_rownames("sample_id")  # adjust column name if different

ps <- phyloseq(
  otu_table(seqtab.nochim, taxa_are_rows = FALSE),
  sample_data(metadata_df),
  tax_table(taxa)
)
ps
```

**`>> YOUR DECISION <<` — choose your engineering variable.**
Look at the columns in `metadata`. Pick ONE variable to use as your grouping
variable for the rest of today's analysis (e.g. plant/process type, country,
or another available metadata field). Write it here:

```{r choose-variable}
grouping_variable <- "process_type"   # <-- replace with your chosen metadata column name
```

---

## 3. Alpha diversity: how rich/even is each community?

```{r alpha-diversity}
plot_richness(ps, x = grouping_variable, measures = c("Observed", "Shannon")) +
  theme_bw()
```

**Question:** Does diversity differ visibly across your grouping variable? Is
that difference large relative to the spread *within* each group?

---

## 4. Beta diversity: how different are communities from each other?

```{r ordination}
ps_rel <- transform_sample_counts(ps, function(x) x / sum(x))  # relative abundance
ord <- ordinate(ps_rel, method = "NMDS", distance = "bray")
```

```{r plot-ordination}
plot_ordination(ps_rel, ord, color = grouping_variable) +
  theme_bw()
```

**Question:** Do samples cluster by your grouping variable, or are they mixed
together? What would clustering (or lack of it) imply about how strongly this
engineering variable shapes the community?

---

## 5. Which taxa differ between groups?

This step only makes sense if your grouping variable has (at least
approximately) two groups. If you have more than two, pick two to compare, or
ask your instructor for a multi-group approach.

```{r top-taxa}
# Collapse to genus level and keep the most abundant genera for a readable plot
ps_genus <- tax_glom(ps_rel, taxrank = "Genus", NArm = FALSE)

top_genera <- names(sort(taxa_sums(ps_genus), decreasing = TRUE))[1:15]
ps_top <- prune_taxa(top_genera, ps_genus)

plot_bar(ps_top, fill = "Genus", x = grouping_variable) +
  theme_bw() +
  labs(y = "Relative abundance")
```

```{r simple-test}
# A simple (not multiple-testing-corrected) group comparison for one genus of interest.
# Replace "GenusOfInterest" with a genus name you saw in the plot above.
genus_of_interest <- "GenusOfInterest"

abund_df <- psmelt(ps_top) %>%
  filter(Genus == genus_of_interest)

wilcox.test(Abundance ~ get(grouping_variable), data = abund_df)
```

**Question:** Is the genus you tested visibly enriched in one group in the bar
plot above? Does the simple statistical test support that impression? (With
this sample size, treat the p-value as a rough signal, not a definitive
answer.)

---

## 6. Produce your deliverable

**Choose ONE figure** from today (alpha diversity, ordination, or the top-taxa
bar plot) that best supports a claim about how the microbial community
relates to your chosen engineering variable.

```{r save-deliverable}
# Re-run/save your chosen plot here, e.g.:
# ggsave(file.path(output_dir, "module1_deliverable.png"), width = 8, height = 5)
```

**Write your interpretation (2–5 sentences) below:**

> *(Replace this line.) State which figure you're using, what pattern it
> shows, and what an engineer/operator running this WWTP process could learn
> or do differently based on it.*

---

## Wrap-up discussion

- Which of today's steps involved a judgment call rather than a fixed rule?
- If you had used a generic 16S reference database instead of MiDAS 4, do you
  think your conclusions today would have changed? Why might a
  purpose-built, ecosystem-specific reference matter here?
