---
title: "Module 1 — Session 1: From raw reads to an ASV table"
subtitle: "27221 Microbiome Engineering — Metabarcoding of wastewater treatment plant communities"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Before you start

Today you will take **real raw 16S rRNA amplicon reads** from wastewater treatment
plants (WWTPs) around the world and turn them into an **ASV table**: a matrix of
exactly which microbial sequence variants were found, in which samples, and how
often.

You are not expected to write R code from scratch today. Every chunk below is
already written. Your job is to **run each chunk, look at what it produces, and
make a small number of justified decisions** — marked `>> YOUR DECISION <<` —
based on the plots and tables you see. There is no single "correct" number for
these decisions; what matters is that your choice is grounded in what the data
show you.

Run chunks one at a time (click the green "Run" arrow, or `Ctrl+Enter` /
`Cmd+Enter`), and read the output before moving to the next chunk.

---

## 0. Setup

```{r libraries}
library(dada2)
library(tidyverse)
```

```{r config}
# --- SESSION CONFIG: adjust these paths if your instructor gives you different ones ---
raw_reads_dir   <- "/path/to/shared/wwtp_16S/raw_reads"     # forward+reverse fastq.gz files
metadata_path   <- "/path/to/shared/wwtp_16S/sample_metadata.csv"
output_dir      <- "~/module1_output"                        # where YOUR results get saved
dir.create(output_dir, showWarnings = FALSE)
```

```{r inspect-inputs}
# Forward and reverse read files are expected to follow the pattern
# SAMPLENAME_R1.fastq.gz / SAMPLENAME_R2.fastq.gz
fnFs <- sort(list.files(raw_reads_dir, pattern = "_R1.fastq.gz$", full.names = TRUE))
fnRs <- sort(list.files(raw_reads_dir, pattern = "_R2.fastq.gz$", full.names = TRUE))
sample.names <- str_remove(basename(fnFs), "_R1.fastq.gz$")

metadata <- read_csv(metadata_path)

# Sanity check: do we have the same number of forward/reverse files, and do
# they match the metadata table?
length(fnFs); length(fnRs)
head(metadata)
```

**Question:** How many samples do you have? Do the sample names in `metadata`
match `sample.names`? (If not, stop here and flag it — everything downstream
depends on this being correct.)

---

## 1. Look at read quality before doing anything else

```{r quality-profiles}
plotQualityProfile(fnFs[1:2])  # forward reads, first two samples
plotQualityProfile(fnRs[1:2])  # reverse reads, first two samples
```

The green line is the median quality score at each position; the orange lines
are the 25th/75th percentiles. Quality typically stays high for a while and
then drops off toward the end of the read — this is normal for Illumina
sequencing, not a problem with your samples.

**`>> YOUR DECISION <<` — truncation length.**
Look at the plots above. At roughly which position does quality drop below a
score of ~30? Note down a sensible truncation length for forward reads
(`truncLenF`) and reverse reads (`truncLenR`) — you want to cut off the
low-quality tail, but forward + reverse reads still need to overlap enough to
merge later (the amplicon region here is short enough that this is not usually
a problem, but keep it in mind).

```{r truncation-decision}
truncLenF <- 0   # <-- replace 0 with your chosen value, based on the plot above
truncLenR <- 0   # <-- replace 0 with your chosen value, based on the plot above
```

---

## 2. Filter and trim

```{r filter-trim}
filtFs <- file.path(output_dir, "filtered", paste0(sample.names, "_F_filt.fastq.gz"))
filtRs <- file.path(output_dir, "filtered", paste0(sample.names, "_R_filt.fastq.gz"))
names(filtFs) <- sample.names
names(filtRs) <- sample.names

out <- filterAndTrim(
  fnFs, filtFs, fnRs, filtRs,
  truncLen = c(truncLenF, truncLenR),
  maxN = 0,        # DADA2 requires no ambiguous bases
  maxEE = c(2, 2),  # max expected errors per read (a stricter/looser choice affects yield vs. quality)
  truncQ = 2,
  rm.phix = TRUE,
  compress = TRUE,
  multithread = TRUE
)
out
```

**Question:** For each sample, what fraction of reads survived filtering
(`reads.out / reads.in`)? Is that fraction roughly consistent across samples,
or are some samples much worse? A sample losing the large majority of its
reads here is worth flagging before you interpret anything from it later.

---

## 3. Learn the error model

DADA2's core idea is to learn a model of how sequencing errors occur at each
position/quality score, so it can tell a true biological variant apart from a
sequencing error.

```{r learn-errors}
errF <- learnErrors(filtFs, multithread = TRUE)
errR <- learnErrors(filtRs, multithread = TRUE)
```

```{r plot-errors}
plotErrors(errF, nominalQ = TRUE)
```

**Question:** The black line is the fitted error model; the points are the
observed error rates. Does the fitted line reasonably follow the observed
points, and do error rates decrease as quality score increases (as expected)?
If the fit looks poor, that's a signal something upstream (e.g. truncation
length) may need revisiting.

---

## 4. Denoise (infer ASVs)

```{r dada}
dadaFs <- dada(filtFs, err = errF, multithread = TRUE)
dadaRs <- dada(filtRs, err = errR, multithread = TRUE)
```

---

## 5. Merge paired reads

```{r merge}
mergers <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose = TRUE)
```

**Question:** What fraction of reads merged successfully, on average? A low
merge rate usually means the truncation lengths chosen in Step 1 left
insufficient overlap between forward and reverse reads.

---

## 6. Build the sequence table

```{r seqtab}
seqtab <- makeSequenceTable(mergers)
dim(seqtab)                       # samples x unique sequence variants
table(nchar(getSequences(seqtab)))  # distribution of sequence lengths
```

**Question:** Do the sequence lengths cluster tightly around the expected
amplicon length for this region, or is there a long tail of unexpectedly
short/long sequences? Unexpected lengths are often a sign of non-specific
amplification and can be filtered out.

---

## 7. Remove chimeras

```{r chimeras}
seqtab.nochim <- removeBimeraDenovo(seqtab, method = "consensus", multithread = TRUE, verbose = TRUE)
dim(seqtab.nochim)
sum(seqtab.nochim) / sum(seqtab)   # fraction of reads retained after chimera removal
```

---

## 8. Track reads through the whole pipeline

This is the single most important sanity-check table in a DADA2 workflow —
always look at it before trusting your results.

```{r track}
getN <- function(x) sum(getUniques(x))
track <- data.frame(
  input      = out[, 1],
  filtered   = out[, 2],
  denoisedF  = sapply(dadaFs, getN),
  denoisedR  = sapply(dadaRs, getN),
  merged     = sapply(mergers, getN),
  nonchim    = rowSums(seqtab.nochim)
)
rownames(track) <- sample.names
track
```

**Question:** Is there any step where a sample loses a disproportionate
fraction of reads compared to the others? If so, which step, and what would
you check next?

---

## 9. Save your output for Session 2

```{r save}
saveRDS(seqtab.nochim, file.path(output_dir, "seqtab_nochim.rds"))
write_csv(metadata, file.path(output_dir, "metadata_used.csv"))
```

---

## Wrap-up discussion

- What does one row of `seqtab.nochim` represent? What does one column
  represent?
- You made a handful of judgment calls today (truncation length, filtering
  stringency). If a classmate had made slightly different choices, would you
  expect wildly different biological conclusions later on? Why or why not?
- Bring your saved `seqtab_nochim.rds` to Session 2 — that's where we turn
  this table into biology.
